Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile

### 2024-01-18 article/paper_drought_Chile2023.tex
actualizar articulo

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  number,
  preprint,
  3p,
  onecolumn]{elsarticle}

@@

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, frame hidden, sharp corners, breakable, borderline west={3pt}{0pt}{shadecolor}, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Precipitation is the primary driver of drought and is intensified by
temperature \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citep{Loon2016} and
\citep{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought. \citep{Crausbay2017}
proposed the ecological drought definition as \emph{``an episodic
deficit in water availability that drives ecosystems beyond thresholds
of vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''}. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} states that even if global warming
is stabilized at 1.5°--2°C, many parts of the world will be impacted by
more severe agricultural and ecological droughts. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

@@

To evaluate meteorological drought (i.e., water supply), the World
Meteorological Organization (WMO; \citep{WMO2012}) recommends the
Standardized Precipitation Index (SPI; \citep{Mckee1993}), a multiscalar
drought index that allows to monitor precipitation deficits from short-
to long-term. Following the same approach, \citep{Vicente-Serrano2010}
incorporates into the SPI the effect of temperature through the use of
potential evapotranspiration, thus proposing the SPEI (Standardized
Precipitation Evapotranspiration Index). Similarly, to evaluate solely
the evaporative demand driven by temperature, \citep{Hobbins2016} and
\citep{McEvoy2016} came up with the Evaporative Demand Drought Index
(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
\citep{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
cumulative Normalized Difference Vegetation Index (NDVI), which could be
acumulated over the growing season or any period (e.g., months),
resulting in a multiscalar drought index. For soil moisture, several
drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
normalized index \citep{Narasimhan2005} and the Soil Moisture
Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
normalized index using vegetation, land surface temperature, and a
vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
estimate the standardized terrestrial water storage index (STI)
\citep{Cui2021}, a standardized anomaly that follows the methodology of
the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
water supply, demand, and storage, which can help to make a
comprehensive assessment of drought.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citep{Beck2023} and the landcover type
persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
The south part of ``Centro'' and the north part of ``Austral'' are
dominated by an Oceanic climate (Cfb). Those zones are high in forest
and grassland. The southern part of the country has a tundra climate,
and in Patagonia, it is a cold semi-arid area with an extended surface
of grassland.

@@

@@

@@

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses SM. Finally, for the proxy
of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
was smoothed following the procedure described in \citep{Zambrano2018}
and \citep{Zambrano2016}.

@@

It corresponds to a moving window (convolution) that sums the variable
for \(s\) starting for the last month \(n\) until the month, which could
sum for \(s\) months (n-s+1). Once the variable is cumulated over time
for the scale \(s\), we used a nonparametric approach following
\citep{Hobbins2016} to derive the drought indices. Thus, the empirically
derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(A_i)\) probabilities across a period of interest:

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses for forest, schrublands, savannas, grasslands, wetlands,
croplands, urban, snow and ice, barren, and water bodies. To validate
the IGBP MODIS, we compare the macroclasses with the ones of a more
detailed landcover map made by \citep{Zhao2016} for Chile for the years
2013--2014 (LCChile). The later has a spatial resolution of 30 m and
three levels of defined classes; from those, we used level 1, which fits
with the macroclasses derived from the IGBP MODIS. We chose the years
2013 and 2014 from MODIS IGBP because the LCChile was made with data
acquired in 2013 and 2014. We resampled LCChile to the spatial
resolution of the IGBP MODIS using the nearest neighbor method. Then, we
took a random sample of 1000 points within continental Chile and
extracted the classes that fell within each point for LCCHILE, IGP2013,
and IGBP2014. Finally, we applied a confusion matrix and estimated the
accuracy and F1 score as metrics of performance. The validation of
IGBP2013 and IGBP2014 with LCChile reached similar metrics of
performance, having an accuracy of 0.82 and a F1 score of 0.66 (see
SS1).

@@

\citep{Vicente-Serrano2021} made a global analysis of the drought's
severity trend using SPI, SPEI, and the Standardized Evapotranspiration
Deficit Index (SEDI; \citep{Vicente-Serrano2018}) to evaluate AED. They
indicate that the increase in hydrological drought has been due to
anthropogenic effects rather than climate change. This is because the
global increase in AED did not explain the change in the spatial pattern
of the hydrological drought. Also, they state that \emph{``the increase
in hydrological droughts has been primarily observed in regions with
high water demand and land cover change''}. We will contrast this
hypothesis with what is occurring in Chile. To achieve this, we will use
the landcover class type that remains more than 80\% of types for
2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
there are low changes.

@@

To assess how water demand and supply and soil moisture are related to
vegetation productivity (zcNDVI), we analyze the linear correlation
between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
36-month time scales against zcNDVI. We followed a similar approach to
that used by \citep{Meroni2016} when using the SPI for meteorological
drought against the cumulative FAPAR (Fraction of Absorbed
Photosynthetically Active Radiation) as a proxy for vegetation
productivity. We made a pixel-to-pixel linear correlation analysis per
index. First, we calculate the Pearson coefficient of correlation for
the six time scales and let the time scale that reaches the maximum
correlation be significant (p \textless{} 0.05). Then, we extracted the
Pearson correlation value corresponding to the time scales that reached
the maximum value. Thus, we derived two raster maps per index, the first
with the time scales and the second with the correlation value.

\hypertarget{validation-of-era5l-variables}{%
\subsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables}}

@@

Figure~\ref{fig-zcNDVI_var} and \ref{tab-landcoverTrend} show that the
macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'', and
``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte Chico''
shows an increase in barrend land of 111 \(km^2 year^{-1}\) and a
reduction in the class savanna -70 \(km^2 year^{-1}\). In the ``Centro''
and ``Sur'' there are changes in the chilean matorral, with an important
reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and an increase in
shrubland and grassland. Showing a change for more dense vegetation
types. It appears to be a change from the surface occupied by cropland
from the ``Centro'' to the ``Sur.'' Also, there is a high increase in
forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the savanna
lost.

\begin{table}[!ht]
\label{tab-landcoverTrend}
\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

@@

Analyzing the water supply in the macrozones ``norte chico,''
``central,'' and ``sur,'' the SPI shows a decreasing trend that
increases at longer time scales due to the prolonged reduction in
precipitation. It reaches a trend of -0.056 (z-score) per decade
(Figure~\ref{fig-trendSPI}) for time scales of 36 months. In the case of
water demand, we analyze the EDDI. It shows a positive trend, caused by
an increase in AED. The trend on EDDI reaches a maximum of 0.053 per
decade in the macrozones ``norte grande'' and ``norte chico''
(\textbf{?@fig-trendEDDI}). The behavior is similar to SPI, having close
trends with opposite signs. Nevertheless, the spatial patterns are
different. The maximum trend for SPI-36 is in ``norte chico'' and ``zona
central.'' For EDDI-36, the maximum trend is from ``northe grande'' to
``zona central.''

@@

\subcaption{\label{fig-trendSPI-1}SPI (Standardized Precipitation
Index)}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

@@

\subcaption{\label{fig-trendSPI-2}SPEI (Standardized Precipitation
Evapotranspiration Index)}
\end{minipage}%
\newline
\begin{minipage}[t]{0.50\linewidth}

@@

\subcaption{\label{fig-trendSPI-3}EDDI (Evaporative Demand Drought
Index)}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

@@

\subcaption{\label{fig-trendSPI-4}SSMI (Standardized Soil Moisture
Index)}
\end{minipage}%

\caption{\label{fig-trendSPI}Linear trend of the drought index (*) at
time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}

@@

\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}{%
\subsection{Impact for water supply and demand, and soil moisture in
vegetation productivity (CHEQUEAR CIFRAS EN EL
TEXTO)}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity-chequear-cifras-en-el-texto}}

According with what is showed Figure~\ref{fig-corTimeScale},
Figure~\ref{fig-corPerson}, and Table 3. Forest seems to be the most
resistant type to drought, showing that only ``Centro'' is slightly (rsq
= 0.3) impacted by a 12-month soil moisture deficit (SSI-12). Following
by grasslands, which show to be impacted by a SSI-12 with a rsq = 0.4,
and a decrease in water supply (SPI-36 and SPEI-24 with rsq = 0.3) in
the ``Norte Chico'' and to a lesser degree in the ``Norte Grande.'' But,
in the southern macrozones, this type was not affected by water demand,
water supply, or soil moisture. The types that show to be most affected
by variation in climate conditions are shrublands, savannas, and
croplands. For savanna in ``Norte Chico,'' the SSI-12 and SPI-24 reached
an rsq of 0.8 and 0.6, respectively. This value decreases to the south,
but the SSI-12 is still the variable explaining more of the variation in
vegetation productivity (rsq = 0.4 in ``Centro'' and 0.2 in ``Sur''). In
the case of croplands, the SPEI-12, SPI-36, and SSI-12 explain between
40\% and 75\% in ``Centro.'' The most affected type by climatic
variation was shrubland, where in ``Norte Chico'' and ``Centro'' soil
moisture and precipitation explained more than 50\%, with SSI-12 being
the most relevant variable, followed by SPI-36 in ``Norte Chico'' and
SPI-24 in ``Sur.''

\begin{table}[!ht]
\label{tab-corLandcover}
\caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36. The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
\end{table}

@@

1.- Respecto a lo que indica \citep{Vicente-Serrano2018}, de que el
aumento en la tendencia en severidad de la sequía (hidrológica) tiene
que ver más con un aumento de la demanda de agua (LULCC) que a una
tendencia en las condiciones climáticas (SPI-12).

@@

5.- Sequía, aridez, water scarcity. Si puede utilizar el mapa y
proyección climática de Koppen\_Geiger \citep{Beck2023} para dirigir
esta discusión

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  authoryear,
  preprint,
  3p,
  onecolumn]{elsarticle}

@@

\usepackage{lineno}\linenumbers \usepackage{multirow} \usepackage{lscape} \newcommand{\blandscape}{\begin{landscape}} \newcommand{\elandscape}{\end{landscape}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\journal{Journal Name}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-harv}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Drought indices of water demand and supply, soil moisture, vegetation, and its impact on LULCC in continental Chile},
  pdfauthor={Francisco Zambrano},
  pdfkeywords={drought, land cover change, satellite},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, sharp corners, breakable, boxrule=0pt]}{\end{tcolorbox}}\fi

@@

Precipitation is the primary driver of drought and is intensified by
temperature \citep{Luo2017}. Drought impacts soil moisture, hydrological
regimes, and vegetation productivity. Initially, drought was commonly
classified as meteorological, hydrological, and agricultural
\citep{Wilhite1985}. Lately, \citet{Loon2016} and
\citet{AghaKouchak2021} have given an updated definition of drought for
the Anthropocene, suggesting that it should be considered the feedback
of humans' decisions and activities that drives the anthropogenic
drought. Even though it has been argued that those definitions do not
fully address the ecological dimensions of drought. \citet{Crausbay2017}
proposed the ecological drought definition as \emph{``an episodic
deficit in water availability that drives ecosystems beyond thresholds
of vulnerability, impacts ecosystem services, and triggers feedback in
natural and/or human systems''}. Moreover, many ecological studies have
misinterpreted how to characterize drought, for example, sometimes
considering ``dry'' conditions as ``drought'' \citep{Slette2019}. On the
other hand, the AR6 \citep{IPCC2023} states that even if global warming
is stabilized at 1.5°--2°C, many parts of the world will be impacted by
more severe agricultural and ecological droughts. Then, there is a
challenge in conducting drought research, especially to evaluate its
impact on ecosystems.

@@

To evaluate meteorological drought (i.e., water supply), the World
Meteorological Organization (WMO; \citet{WMO2012}) recommends the
Standardized Precipitation Index (SPI; \citet{Mckee1993}), a multiscalar
drought index that allows to monitor precipitation deficits from short-
to long-term. Following the same approach, \citet{Vicente-Serrano2010}
incorporates into the SPI the effect of temperature through the use of
potential evapotranspiration, thus proposing the SPEI (Standardized
Precipitation Evapotranspiration Index). Similarly, to evaluate solely
the evaporative demand driven by temperature, \citet{Hobbins2016} and
\citet{McEvoy2016} came up with the Evaporative Demand Drought Index
(EDDI). For vegetation, in a similar manner as the SPI, SPEI and EDDI;
\citet{Zambrano2018} proposed the zcNDVI, a standardized anomaly of the
cumulative Normalized Difference Vegetation Index (NDVI), which could be
acumulated over the growing season or any period (e.g., months),
resulting in a multiscalar drought index. For soil moisture, several
drought indices exist, such as the Soil Moisture Deficit Index (SDMI) a
normalized index \citep{Narasimhan2005} and the Soil Moisture
Agricultural Drought Index (SMADI) \citep{Souza2021} which is a
normalized index using vegetation, land surface temperature, and a
vegetation condition index (VCI, \citep{Kogan1995}). From TWS, we can
estimate the standardized terrestrial water storage index (STI)
\citep{Cui2021}, a standardized anomaly that follows the methodology of
the SPI, SPEI, EDDI, and zcNDVI. Thereby, we have drought indices for
water supply, demand, and storage, which can help to make a
comprehensive assessment of drought.

@@

Continetal Chile has a diverse climate condition from north to south and
east to west \citep{Aceituno2021} (Figure~\ref{fig-studyArea}), which
determines its great ecosystem diversity (Figure~\ref{fig-LCprop}). The
Andes Mountains are a main factor in latitudinal variation
\citep{Garreaud2009}. To describe the climate and ecosystem of Chile, we
use the Koppen-Geiger release by \citet{Beck2023} and the landcover type
persistance of 80\% of times for 2001--2022, from the IGBP
classification scheme \citep{Friedl2019} (see
Section~\ref{sec-methods_lulc}). ``Norte Grande'' and ``Norte Chico''
predominate in an arid desert climate with hot (Bwh) and cold (Bwk)
temperatures. At the south of ``Norte Chico,'' the climate changes to an
arid steppe with cold temperatures (Bsk). Mainly, the land is barren,
with a minor surface of vegetation types such as shrubland and
grassland. In the zones ``Centro'' and the north half of ``Sur,'' the
main climate is Mediterranean, with warmer to hot summers (Csa and Csb).
In ``Centro,'' there is a major proportion (\textasciitilde50\%) of
Chilean matorral (shrubland and savanna, \citep{Fuentes2021}), followed
by grassland (\textasciitilde16\%), forest (8\%), and croplands (5\%).
The south part of ``Sur'' and the north part of ``Austral'' are
dominated by an Oceanic climate (Cfb). Those zones are high in forest
and grassland. The southern part of the country has a tundra climate,
and in Patagonia, it is a cold semi-arid area with an extended surface
of grassland, forest, and, to a lesser extent, savanna.

@@

\hypertarget{validation-of-era5l-variables}{%
\subsubsection{Validation of ERA5L
variables}\label{validation-of-era5l-variables}}

@@

\hypertarget{atmospheric-evaporative-demand-aed}{%
\subsubsection{Atmospheric Evaporative Demand
(AED)}\label{atmospheric-evaporative-demand-aed}}

@@

\hypertarget{non-parametric-calculation-of-drought-indices}{%
\subsubsection{Non-parametric calculation of drought
indices}\label{non-parametric-calculation-of-drought-indices}}

To evaluate water demand, we chose the \(EDDI\)
\citep{Hobbins2016, McEvoy2016} index, which uses the \(AED\). For
supply, we used the index recommended by the World Meteorological
Organization (WMO) for monitoring drought, the SPI \citep{Mckee1993}. We
calculated the SPEI, which used a balance between \(P\) and \(AED\), in
this case, an auxiliary variable \(D = P-AED\) is used. In this study,
we used the \(SSI\) (standardized soil moisture index at 1 m)
\citep{Hao2013, AghaKouchak2014}, which uses SM. Finally, for the proxy
of productivity, \(zcNDVI\), we used the NDVI. Before using the NDVI, it
was smoothed using a locally-weighted polynomial regression, following
the procedure described in \citet{Zambrano2018} and
\citet{Zambrano2016}.

@@

It corresponds to a moving window (convolution) that sums the variable
for \(s\) starting for the last month \(n\) until the month, which could
sum for \(s\) months (n-s+1). Once the variable is cumulated over time
for the scale \(s\), we used a nonparametric approach following
\citet{Hobbins2016} to derive the drought indices. Thus, the empirically
derived probabilities are obtained through an inverse normal
approximation \citep{Abramowitz1968}. Then, we used the empirical Tukey
plotting position \citep{Wilks2011} over \(A_i\) to derive the
\(P(A_i)\) probabilities across a period of interest:

@@

To analyze the LULCC, we use the IGBP scheme from the MCD12Q1 collection
6.1 from MODIS. This product has a yearly frequency from 2001 to 2022.
The IGBP defines 17 classes; from these, we regrouped into ten
macroclasses for forest, schrublands, savannas, grasslands, wetlands,
croplands, urban, snow and ice, barren, and water bodies. To validate
the IGBP MODIS, we compare the macroclasses with the ones of a more
detailed landcover map made by \citet{Zhao2016} for Chile for the years
2013--2014 (LCChile). The later has a spatial resolution of 30 m and
three levels of defined classes; from those, we used level 1, which fits
with the macroclasses derived from the IGBP MODIS. We chose the years
2013 and 2014 from MODIS IGBP because the LCChile was made with data
acquired in 2013 and 2014. We resampled LCChile to the spatial
resolution of the IGBP MODIS using the nearest neighbor method. Then, we
took a random sample of 1000 points within continental Chile and
extracted the classes that fell within each point for LCCHILE, IGP2013,
and IGBP2014. Finally, we applied a confusion matrix and estimated the
accuracy and F1 score as metrics of performance. The validation of
IGBP2013 and IGBP2014 with LCChile reached similar metrics of
performance, having an accuracy of 0.82 and a F1 score of 0.66 (see
SS1).

@@

\citet{Vicente-Serrano2021} made a global analysis of the drought's
severity trend using SPI, SPEI, and the Standardized Evapotranspiration
Deficit Index (SEDI; \citet{Vicente-Serrano2018}) to evaluate AED. They
indicate that the increase in hydrological drought has been due to
anthropogenic effects rather than climate change. This is because the
global increase in AED did not explain the change in the spatial pattern
of the hydrological drought. Also, they state that \emph{``the increase
in hydrological droughts has been primarily observed in regions with
high water demand and land cover change''}. We will contrast this
hypothesis with what is occurring in Chile. To achieve this, we will use
the landcover class type that remains more than 80\% of types for
2001--2022 to evaluate the trend on zcNDVI and use this as a mask where
there are low changes.

@@

To assess how water demand and supply and soil moisture are related to
vegetation productivity (zcNDVI), we analyze the linear correlation
between the indices SPI, SPEI, EDDI, and SSI for 1, 3, 6, 12, 24, and
36-month time scales against zcNDVI. We followed a similar approach to
that used by \citet{Meroni2016} when using the SPI for meteorological
drought against the cumulative FAPAR (Fraction of Absorbed
Photosynthetically Active Radiation) as a proxy for vegetation
productivity. We made a pixel-to-pixel linear correlation analysis per
index. First, we calculate the Pearson coefficient of correlation for
the six time scales and let the time scale that reaches the maximum
correlation be significant (p \textless{} 0.05). Then, we extracted the
Pearson correlation value corresponding to the time scales that reached
the maximum value. Thus, we derived two raster maps per index, the first
with the time scales and the second with the correlation value.

@@

Figure~\ref{fig-zcNDVI_var} and Table \ref{tab-landcoverTrend} show that
the macrozones with major LULCC for 2001-2022 were ``Centro'', ``Sur'',
and ``Austral'' with 36\%, 31\%, and 34\%, respectively. The ``Norte
Chico'' shows an increase in barrend land of 111 \(km^2 year^{-1}\) and
a reduction in the class savanna -70 \(km^2 year^{-1}\). In the
``Centro'' and ``Sur'' there are changes in the Chilean matorral, with
an important reduction in savanna (-136 to -318 \(km^2 yr^{-1}\)), and
an increase in shrubland and grassland. Showing a change for more dense
vegetation types. It appears to be a change from the surface occupied by
cropland from the ``Centro'' to the ``Sur.'' Also, there is a high
increase in forest (397 \(km^2 yr^{-1}\)) in the ``Sur,'' replacing the
savanna lost.

\begin{table}[!ht]
\caption{The value of Sen's slope trend next to the time-series plot of surface per landcover class (IGBP MCD12Q1.016) for 2001–2022 through Central Chile. Values of zero indicate that there was not a significant trend. Red dots on the plots indicate the maximum and minimum values of surface.}
\label{tab-landcoverTrend}
\includegraphics[]{../output/figs/table_var_landcover_macro.png}
\end{table}

@@

Analyzing the water supply, the macrozones that have the lowest trend
are ``Norte Chico'' and ``Centro,'' where the SPI, SPEI, and SSI show
that it decreases at longer time scales due to the prolonged reduction
in precipitation. At 36 months, it reaches trends between -0.03 and
-0.04 (z-score) per decade for SPI, SPEI, and SSI
(Figure~\ref{fig-trendDI}). For ``Sur,'' the behavior is similar,
decreasing at longer scales and having between -0.016 and -0.025 per
decade for SPI, SPEI, and SSI. On the other hand, all macrozones show an
increase in the trend in all the drought indices, with ``Norte Grande''
having the highest at 36 months (0.042 per decade). Because of this, the
SPEI (which uses AED) reached its lowest value in ``Norte Grande,'' with
-0.03 at 36 months. Despite the other macrozones, ``Austral'' showed an
increase in all indices, being the highest for EDDI at 36 months (0.025)
and the lowest for SSI, which shows only a minor increase in the trend
(Figure~\ref{fig-trendDI}).

\hypertarget{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}{%
\subsection{Impact for water supply and demand, and soil moisture in
vegetation
productivity}\label{impact-for-water-supply-and-demand-and-soil-moisture-in-vegetation-productivity}}

According to what is shown in Figure~\ref{fig-corTimeScale},
Figure~\ref{fig-corPerson}, and Table \ref{tab-corLandcover}, forest
seems to be the most resistant type to drought. Showing that only
``Centro'' is slightly (rsq = 0.25) impacted by a 12-month soil moisture
deficit (SSI-12). Grasslands are shown to be impacted by a SSI-12 with a
rsq = 0.45 and by a decrease in water supply (SPI-36 and SPEI-24 with
rsq = 0.28 and 0.34, respectively) in the ``Norte Chico'' and to a
lesser degree in the ``Norte Grande.'' But, in further south macrozones,
this type was not affected by water demand, water supply, or soil
moisture. The types that show to be most affected by variation in
climate conditions are shrublands, savannas, and croplands. For savannas
in ``Norte Chico,'' the SSI-12 and SPI-24 reached an rsq of 0.74 and
0.58, respectively. This value decreases to the south, but the SSI-12 is
still the variable explaining more of the variation in vegetation
productivity (rsq = 0.45 in ``Centro'' and 0.2 in ``Sur''). In the case
of croplands, the SPEI-12, SPI-36, and SSI-12 explain between 45\% and
66\% of ``Norte Chico.'' The most affected type by climatic variation
was shrubland, where in ``Norte Chico'' and ``Centro'' soil moisture
explained 59\% and precipitation, 37\%, with SSI-12 being the most
relevant variable, followed by SPI-36 in ``Norte Chico'' and SPI-24 in
``Sur.''

@@

\subcaption{\label{fig-trendDI-1}SPI (Standardized Precipitation Index)}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

@@

\subcaption{\label{fig-trendDI-2}SPEI (Standardized Precipitation
Evapotranspiration Index)}
\end{minipage}%
\newline
\begin{minipage}[t]{0.50\linewidth}

@@

\subcaption{\label{fig-trendDI-3}EDDI (Evaporative Demand Drought
Index)}
\end{minipage}%
%
\begin{minipage}[t]{0.50\linewidth}

@@

\subcaption{\label{fig-trendDI-4}SSMI (Standardized Soil Moisture
Index)}
\end{minipage}%

\caption{\label{fig-trendDI}Linear trend of the drought index (*) at
time scales of 1, 3, 6, 12, 24, and 36 months for 1981-2023}

@@

\begin{table}[!ht]
\caption{Summarry per landcover macroclass and macrozone regarding the correlation between zcNDVI with the drought indices EDDI, SPI, SPEI, and SSI for time scales of 1, 3, 6, 12, 24, and 36. The numbers in each cell indicate the time scale that reached the maximum correlation for the landcover and macrozone, and the color indicates the strength of the r-squared obtained with the index and the time scale.}
\label{tab-corLandcover}
\includegraphics[]{../output/figs/tabla_r_cor_macro_indice.png}
\end{table}

@@

1.- Respecto a lo que indica \citet{Vicente-Serrano2018}, de que el
aumento en la tendencia en severidad de la sequía (hidrológica) tiene
que ver más con un aumento de la demanda de agua (LULCC) que a una
tendencia en las condiciones climáticas (SPI-12).

@@

5.- Sequía, aridez, water scarcity \citep{VanLoon2013} Si puede utilizar
el mapa y proyección climática de Koppen\_Geiger \citep{Beck2023} para
dirigir esta discusión
